\documentclass[11pt]{article}

\usepackage[margin=0.82in]{geometry}
\usepackage{amsmath,amssymb,booktabs,tabularx,array,xcolor,float}
\usepackage[round]{natbib}
\usepackage[colorlinks=true,citecolor=blue!45!black,urlcolor=blue!45!black]{hyperref}
\usepackage{microtype}

\setlength{\parindent}{0pt}
\setlength{\parskip}{0.42em}
\setlength{\emergencystretch}{3em}

\title{Calibration of the Quantitative Model\\
\large Provisional update}
\author{Tommaso De Santo}
\date{}

\begin{document}
\maketitle

\section{Calibration}
\label{sec:quant_calibration}

I organize the model parameters into two groups. The first consists of
parameters assigned from external evidence, measured directly in the data, or
fixed as maintained restrictions. The second consists of parameters estimated
internally by matching model moments to their empirical counterparts. I describe
the two groups in turn, explain which moments are particularly informative
about the internally estimated parameters, and then report the model's fit.

\subsection{Parameters Set Outside the Estimation}

I first describe the parameters assigned outside the estimation.
Table~\ref{tab:quant_assigned} summarizes their values and sources.

\paragraph{Demographics and preferences.}
A model period is four years. Households enter adult life at age 18, retire at
age 66, and live at most through age 85. Survival is certain before retirement
and thereafter follows four-year probabilities constructed from the combined
male and female survivor counts in the 2023 Social Security period life table.
Dependent children remain in the household for an expected 18 years, implying
a maturity probability of $2/9$ per period. I set risk aversion to $\sigma=2$.

\paragraph{Housing finance and transaction costs.}
Following \citet{greaneyDynamicUrbanEconomics2025}, I set the annual real
return on liquid wealth to $i_{\mathrm{ann}}=0.02$, annual housing depreciation
to $\delta_{\mathrm{ann}}=0.011$, the proportional selling cost to
$\psi^s=0.06$, and the labor-income tax rate to
$\tau_{\mathrm{pay}}=0.179$. The return is based on the average ten-year real
interest rate, depreciation combines BEA residential-structure depreciation
with an adjustment for the land share of house values, and the tax rate comes
from OECD data. I set the annual property-tax rate to
$\tau^p_{\mathrm{ann}}=0.01$. Mortgage debt cannot exceed a share $\phi=0.80$
of the value of the house, so a purchase requires at least 20 percent equity.
The benchmark imposes no separate payment-to-income limit.

\paragraph{Income process.}
I impose a deterministic lifecycle profile of household earnings, $e_a$,
normalized to average one over working ages. Within age, log income follows an
annual AR(1) process with persistence $\rho_z=0.960$, which I sample at the
model's four-year frequency and approximate with a five-state Rouwenhorst
chain. This persistence is near the upper end of the PSID-based estimates
summarized by \citet{sommerImplicationsUSTax2018}. I set the annual innovation
standard deviation to $\sigma_z=0.0645$, below estimates of gross income risk
in that literature. I treat this variance as provisional rather than as an
externally validated estimate of income risk.\footnote{With innovation risk in
the empirical range, households in the lowest persistent-income states cannot
always finance the model's hard Stone--Geary minimum bundle. The current model
omits insurance margins that could resolve this feasibility conflict.}
The quantitative results are conditional on this restriction.

\paragraph{Housing products and supply.}
Housing is indivisible and available in a limited assortment of types and
sizes \citep{davisHousingFinanceMacroeconomy2015}. I represent this lumpiness
with five owner-occupied house sizes,
$\{H_k\}_{k=1}^5=\{2,4,6,8,10\}$ rooms. Renters choose housing continuously up
to $\bar h^R=6$ rooms, the 90th percentile of the rental-room distribution in
the matched ACS sample. These menus summarize the observed tenure-specific
distribution of rooms. I set the housing-supply elasticity to $\eta=1.75$,
following the population-weighted metropolitan estimate in
\citet{saizGeographicDeterminantsHousing2010}.

\paragraph{Entry wealth and bequests.}
Households enter at age 18 with a draw from the PSID net-worth-to-income
distribution of childless renters aged 18--24. I impose $\theta_n=0$, so
bequest utility does not vary directly with the number of children. Children
can still affect estate accumulation through their effects on housing needs,
tenure, and saving.

\begin{table}[H]
\centering
\caption{Parameters set outside the estimation}
\label{tab:quant_assigned}
\begingroup
\small
\renewcommand{\arraystretch}{1.04}
\begin{tabularx}{\textwidth}{@{}
>{\raggedright\arraybackslash}p{0.31\textwidth}
>{\raggedright\arraybackslash}p{0.21\textwidth}
>{\raggedright\arraybackslash}X@{}}
\toprule
Parameter & Value & Source or restriction \\
\midrule
Model period & 4 years & Model timing \\
Entry, retirement, and maximum ages & 18, 66, and 85 & Model timing \\
Post-retirement survival & Age-specific & 2023 Social Security period life table \\
Expected years with dependent children & 18 years & Child-maturity restriction \\
Risk aversion $\sigma$ & 2 & Maintained curvature \\
Annual real return $i_{\mathrm{ann}}$ & 2.0\% &
Average 10-year real rate, 1962--2024; \citet{greaneyDynamicUrbanEconomics2025} \\
Annual housing depreciation $\delta_{\mathrm{ann}}$ & 1.1\% &
BEA structures depreciation, adjusted for the residential land share;
\citet{greaneyDynamicUrbanEconomics2025} \\
Annual property tax $\tau^p_{\mathrm{ann}}$ & 1.0\% & Maintained tax rate \\
Financed share $\phi$ & 0.80 &
20\% down-payment restriction; \citet{davisHousingFinanceMacroeconomy2015} \\
Sale cost $\psi^s$ & 6\% &
\citet{davisHousingFinanceMacroeconomy2015,greaneyDynamicUrbanEconomics2025} \\
Labor-income tax $\tau_{\mathrm{pay}}$ & 17.9\% &
OECD U.S. average; \citet{greaneyDynamicUrbanEconomics2025} \\
Income persistence $\rho_z$ & 0.960 annually & See text \\
Income innovation standard deviation $\sigma_z$ & 0.0645 annually & See text \\
Owner house sizes $\{H_k\}$ & $\{2,4,6,8,10\}$ rooms &
Occupied-room distribution in the ACS; see text \\
Largest rental unit $\bar h^R$ & 6 rooms & Matched ACS rental-size distribution \\
Housing-supply elasticity $\eta$ & 1.75 &
\citet{saizGeographicDeterminantsHousing2010} \\
Entrant wealth distribution & Wealth-to-income distribution &
Childless renters aged 18--24 in the PSID \\
Child shifter in bequest utility $\theta_n$ & 0 & Maintained restriction \\
\bottomrule
\end{tabularx}
\endgroup
\end{table}

\subsection{Internally Estimated Parameters}

I estimate the remaining 14 parameters using 14 moments describing saving and
bequests, baseline consumption and housing demand, child costs and family size,
tenure, and aggregate housing. The parameters are chosen jointly. The mapping
is nevertheless structured in low-dimensional blocks: a three-parameter
saving--bequest block, a three-parameter baseline demand block, a
two-parameter fertility block, and six parameters with distinct primary
counterparts. Table~\ref{tab:quant_parameters} reports the estimates, search
regions, and the primary empirical counterpart of each parameter.

\begin{table}[H]
\centering
\caption{Internally estimated parameters and primary empirical counterparts}
\label{tab:quant_parameters}
\scriptsize
\renewcommand{\arraystretch}{1.05}
\begin{tabularx}{\textwidth}{@{}l r c c >{\raggedright\arraybackslash}X@{}}
\toprule
Parameter & Estimate & Search region & Near bound & Primary empirical counterpart \\
\midrule
$\beta_{\mathrm{annual}}$ & 0.9991 & $[0.800,0.9995]$ & Yes &
Aggregate wealth / annual gross labor earnings \\
$\theta_0$ & 0.0460 & $[0,8]$ & Yes &
Annual bequest flow / aggregate wealth \\
$\theta_1$ & 0.0944 & $[0.02,16]$ & Yes &
Late-life total-wealth p90/p50 \\
\addlinespace
$\alpha_c$ & 0.6145 & $[0.02,0.98]$ & No &
Childless-renter rent-expenditure slope \\
$\bar c_0$ & 0.5090 & $[0,2]$ & No &
Childless-renter intercept at the mean rent per room \\
$\bar h_0$ & 0.5139 & $[0.05,5.8]$ & No &
Mean rooms among bottom-quintile childless renters \\
\addlinespace
$\bar c_n$ & 0.1553 & $[0,3]$ & No &
Family-size coefficient in the consumption auxiliary \\
$\bar h_{\mathrm{jump}}$ & 0.0522 & $[0,8]$ & Yes &
Housing increment associated with the first child \\
$\bar h_n$ & 0.9258 & $[0,5]$ & No &
Rooms gap: 3+ versus 1--2 children \\
$\psi_{\mathrm{child}}$ & $-1.0261$ & $[-3,3]$ & No &
Completed fertility, jointly with childlessness \\
$\kappa_F$ & 5.4615 & $[0.02,50]$ & No &
Childlessness, jointly with completed fertility \\
\addlinespace
$\chi_O$ & 1.0745 & $[0.1,5]$ & No &
Ownership rate, ages 30--55 \\
$\kappa_T$ & 0.0190 & $[0,0.12]$ & No &
Four-year residual tenure Brier score \\
$\bar H$ & 4.9399 & $[0.2,80]$ & No &
Mean occupied rooms, ages 18--85 \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Saving.}
The annual discount factor governs households' willingness to carry resources
across ages. Its primary target is aggregate household net worth divided by
annual gross labor earnings, measured in the 2005--2019 PSID. The model
constructs the same ratio: beginning-of-period liquid wealth plus gross housing
value in the numerator, and gross labor earnings of working households in the
denominator. Mortgage debt is already contained in liquid wealth. The target is
$6.873$; this measured gross/gross ratio replaces the previously borrowed
$6.9$ ratio with an after-tax denominator. Age-binned wealth ratios remain
untargeted lifecycle diagnostics.

\paragraph{Housing demand, tenure, and supply.}
For a childless renter with allocatable expenditure $x=c+rh$, the Stone--Geary
first-order condition implies
\begin{equation}
rh=-(1-\alpha_c)\bar c_0+(1-\alpha_c)x+\alpha_c r\bar h_0.
\label{eq:les_mapping}
\end{equation}
This creates a triangular three-parameter block. The rent-expenditure slope is
most informative about $\alpha_c$; the intercept at the sample's mean rent per
room combines $\bar c_0$ and $\bar h_0$; and mean rooms among childless renters
in the bottom expenditure quintile pins down the housing level and separates
the two minimum requirements. The same sample restrictions, trimming rule, and
auxiliary regressions are applied to the CEX and simulated model data. The
parameters are therefore estimated jointly from three moments; none of the
three empirical coefficients is itself a structural parameter.

The owner-service premium $\chi_O$ is primarily disciplined by the ownership
rate at ages 30--55. The tenure dispersion $\kappa_T$ is disciplined by a
four-year Brier score: the mean squared error in predicting ownership four
years ahead using only state variables that can be constructed in both the
PSID and the model. This auxiliary moment isolates residual tenure variation
more directly than the unconditional switching rate. Finally, the supply scale
$\bar H$ is disciplined by mean occupied rooms at ages 18--85.

\paragraph{Consumption needs and family size.}
The consumption requirement is $\bar c_0+\bar c_n n_s$. The empirical
counterpart for $\bar c_n$ is the coefficient on family size in a consumption
regression that also controls for rooms, income, age, and ownership; the
identical regression is run on the simulated population. Thus the target
$0.0752$ is an auxiliary coefficient that disciplines $\bar c_n$, not the
numerical value of $\bar c_n$ itself. The housing requirement is
$\bar h_0+\bar h_{\mathrm{jump}}\mathbf 1\{n_s\geq1\}+\bar h_n n_s$. The
first-child rooms increment is most informative about
$\bar h_{\mathrm{jump}}$, while the rooms difference between households with
three or more children and those with one or two is most informative about
$\bar h_n$. These are also model--data moment comparisons, rather than
identities between reduced-form coefficients and primitives.

Completed fertility and childlessness form a two-parameter, two-moment block
for $(\psi_{\mathrm{child}},\kappa_F)$. The average utility value of children
primarily shifts the level of completed fertility, while the choice-dispersion
parameter primarily governs how much mass remains childless. Because both
parameters move both moments, the pair is estimated jointly.

\paragraph{Bequests.}
Bequeathable wealth is liquid wealth plus gross owner-occupied housing. The
child-blind warm-glow function is
\[
\mathcal B(W)=
\theta_0\frac{(\theta_1+\max\{W,0\})^{1-\sigma}
-\theta_1^{1-\sigma}}{1-\sigma},
\]
so $\theta_0$ controls the overall strength of the bequest motive and
$\theta_1$ controls its wealth dependence. The annual bequest-flow-to-wealth
ratio is most informative about $\theta_0$, while the p90/p50 ratio of total
wealth among living PSID households aged 76--84 is most informative about
$\theta_1$. Cross-sectional wealth is measured at the beginning of the model
period; bequests are measured after the saving decision at the death node.
Together with the aggregate wealth ratio, these moments form a joint
three-parameter block for $(\beta,\theta_0,\theta_1)$, following the broader
strategy of separating lifecycle saving, bequest intensity, and wealth
dispersion in \citet{denardiYang2014}.

\subsection{Target Moments and Fit}

The target system uses equal squared percentage deviations: the weight on
moment $m$ is $1/m_{\mathrm{data}}^2$. This transparent preliminary objective
puts the moments in comparable relative units; it is not an estimated optimal
GMM weighting matrix. Table~\ref{tab:quant_fit} reports the complete target
system and its strict model counterpart.

\begin{table}[H]
\centering
\caption{Target moments and model fit}
\label{tab:quant_fit}
\scriptsize
\renewcommand{\arraystretch}{1.03}
\begin{tabularx}{\textwidth}{@{}>{\raggedright\arraybackslash}X r r r r r@{}}
\toprule
Moment & Data & Model & Gap & Weight & Loss \\
\midrule
\multicolumn{6}{@{}l}{\emph{Saving and bequests}} \\
Wealth / annual gross labor earnings & 6.8731 & 5.1365 & $-1.7366$ & 0.0212 & 0.06384 \\
Annual bequest flow / aggregate wealth & 0.0088 & 0.0081 & $-0.0007$ & 12913.22 & 0.00711 \\
Total-wealth p90/p50, ages 76--84 & 3.4481 & 1.8926 & $-1.5556$ & 0.0841 & 0.20352 \\
\addlinespace
\multicolumn{6}{@{}l}{\emph{Baseline consumption and housing demand}} \\
Childless-renter rent-expenditure slope & 0.2647 & 0.2831 & 0.0185 & 14.2734 & 0.00486 \\
Childless-renter intercept at mean price & 0.1064 & 0.1068 & 0.0004 & 88.3061 & 0.00001 \\
Bottom-quintile childless-renter mean rooms & 3.8974 & 3.8348 & $-0.0627$ & 0.0658 & 0.00026 \\
\addlinespace
\multicolumn{6}{@{}l}{\emph{Family size and family housing}} \\
Family-size consumption coefficient & 0.0752 & 0.0752 & $-0.0000$ & 176.8312 & 0.00000 \\
Housing increment with the first child, rooms & 0.6644 & 0.6653 & 0.0009 & 2.2651 & 0.00000 \\
Rooms gap, 3+ versus 1--2 children & 0.3677 & 0.3640 & $-0.0037$ & 7.3963 & 0.00010 \\
Completed fertility & 1.9180 & 2.0854 & 0.1674 & 0.2718 & 0.00761 \\
Childlessness rate & 0.1880 & 0.1925 & 0.0045 & 28.2933 & 0.00057 \\
\addlinespace
\multicolumn{6}{@{}l}{\emph{Tenure and aggregate housing}} \\
Ownership rate, ages 30--55 & 0.5755 & 0.5750 & $-0.0005$ & 3.0196 & 0.00000 \\
Four-year residual tenure Brier score & 0.1176 & 0.1177 & 0.0001 & 72.2923 & 0.00000 \\
Mean occupied rooms, ages 18--85 & 5.7800 & 5.6876 & $-0.0924$ & 0.0299 & 0.00026 \\
\midrule
\multicolumn{5}{@{}r}{Total loss} & 0.28814 \\
\bottomrule
\end{tabularx}
\end{table}

The calibration matches the baseline demand auxiliaries, childlessness, the
family housing moments, ownership, tenure dispersion, and aggregate rooms
closely. Completed fertility is $2.085$, compared with $1.918$ in the data.
The central failure is the wealth block: the model produces a wealth-to-gross-
earnings ratio of $5.136$, compared with $6.873$, and a late-life wealth
p90/p50 ratio of $1.893$, compared with $3.448$. These two rows account for
92.8 percent of the objective.

The annual discount factor moves to $0.9991$, close to its upper search bound
of $0.9995$, yet the model still underproduces both aggregate wealth and
late-life dispersion. A conditional profile is nearly flat at the top:
reoptimizing the other parameters at annual discount factors $0.9990$ and
$0.9995$ gives losses $0.2956$ and $0.2960$. The estimate should therefore not
be interpreted as a precise direct estimate of patience. The present
specification selects a very patient region because it lacks enough wealth
accumulation and dispersion even there.

The calibration remains provisional. In particular, the income-risk process
is constrained by Stone--Geary feasibility, the wealth block points to missing
late-life heterogeneity, and four parameters are near a search boundary.
These limitations should be reported alongside the current fit rather than
hidden by a one-to-one identification claim.

\begin{thebibliography}{9}

\bibitem[Davis and Van Nieuwerburgh(2015)]{davisHousingFinanceMacroeconomy2015}
Davis, M. A., and S. Van Nieuwerburgh. 2015.
``Housing, Finance, and the Macroeconomy.''
\emph{Handbook of Regional and Urban Economics} 5: 753--811.

\bibitem[De Nardi and Yang(2014)]{denardiYang2014}
De Nardi, M., and F. Yang. 2014.
``Bequests and Heterogeneity in Retirement Wealth.''
\emph{European Economic Review} 72: 182--196.

\bibitem[Greaney et al.(2025)]{greaneyDynamicUrbanEconomics2025}
Greaney, B., A. Parkhomenko, and S. Van Nieuwerburgh. 2025.
``Dynamic Urban Economics.'' Working paper.

\bibitem[Saiz(2010)]{saizGeographicDeterminantsHousing2010}
Saiz, A. 2010.
``The Geographic Determinants of Housing Supply.''
\emph{Quarterly Journal of Economics} 125(3): 1253--1296.

\bibitem[Sommer and Sullivan(2018)]{sommerImplicationsUSTax2018}
Sommer, K., and P. Sullivan. 2018.
``Implications of US Tax Policy for House Prices, Rents, and Homeownership.''
\emph{American Economic Review} 108(2): 241--274.

\end{thebibliography}

\end{document}
