\documentclass[10pt]{article}
\usepackage[margin=0.55in]{geometry}
\usepackage{booktabs}
\usepackage{array}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{xcolor}
\setlength{\parindent}{0pt}
\setlength{\parskip}{3pt}
\setlength{\tabcolsep}{4pt}
\renewcommand{\arraystretch}{1.05}

\begin{document}

\begin{center}
{\Large Intergenerational Housing-Fertility Calibration Note}\\
{\normalsize June 23, 2026. Diagnostic status, not a production calibration.}
\end{center}

\textbf{Active object.}
The current quantitative object is the one-market, no-location, Markov-income
intergenerational model in \texttt{code/model/intergen\_housing\_fertility/}.
The old center-periphery \texttt{dt\_cp\_model} code is not the active
calibration object for this note. The relevant recent evidence is the full
global room-gap search and the 14-moment rescore, both documented under
\texttt{docs/model/}.

\textbf{Parameter vector.}
The search varies 13 parameters:
\[
\theta=\{\beta,\alpha,b_0,\bar c_0,\bar c_n,\bar h_0,
\bar h_{\mathrm{jump}},\bar h_n,\psi_{\mathrm{child}},
\kappa_n,\chi,\theta_0,\theta_n\}.
\]
Here \(\beta\) controls saving and old wealth, \(\alpha\) the consumption-space
tradeoff, \(b_0\) entry wealth/access, \((\bar c_0,\bar c_n)\) child consumption
needs, \((\bar h_0,\bar h_{\mathrm{jump}},\bar h_n)\) housing needs,
\(\psi_{\mathrm{child}}\) and \(\kappa_n\) fertility level/dispersion,
\(\chi\) the owner-service premium, and \((\theta_0,\theta_n)\) bequest level
and fertility scaling.

\textbf{Preferred diagnostic objective.}
The previous full room-gap target system had 17 hard moments for 13 parameters.
For interpretation, the cleaner diagnostic objective is 14 hard moments:
one more moment than parameters. The three demoted moments remain diagnostics,
not discarded information.

\begin{center}
\scriptsize
\begin{tabular}{p{2.65in}rrrrp{1.55in}}
\toprule
Moment & Target & Model & Gap & Weight & Role \\
\midrule
TFR & 1.700 & 1.764 & 0.064 & 20.0 & Fertility level \\
Childless rate & 0.150 & 0.276 & 0.126 & 20.0 & Fertility extensive margin \\
Overall ownership & 0.575 & 0.490 & -0.085 & 100.0 & Tenure level \\
Parent-child ownership gap & 0.168 & 0.291 & 0.123 & 45.0 & Family tenure wedge \\
Housing increment \(0\to1\) child & 0.664 & 1.110 & 0.446 & 14.0 & First-child housing demand \\
Housing increment \(1\to2\) children & 0.488 & 0.331 & -0.157 & 8.0 & Marginal child housing demand \\
Young liquid wealth/income & 0.179 & 0.593 & 0.414 & 12.0 & Access and saving \\
Old nonhousing wealth/income median & 2.230 & 2.059 & -0.172 & 0.8 & Lifecycle saving \\
Old parent-childless NH wealth gap & 1.007 & 0.291 & -0.717 & 2.0 & Bequest/composition \\
Childless renter mean rooms & 3.805 & 4.560 & 0.755 & 6.0 & Renter space level \\
Owner-renter room gap, childless & 2.419 & 1.493 & -0.926 & 12.0 & Space separation \\
Childless owner share rooms \(\ge 6\) & 0.596 & 0.762 & 0.165 & 25.0 & Owner large-home use \\
Old ownership & 0.764 & 0.877 & 0.113 & 160.0 & Old retention/exit \\
Young ownership, ages 25--34 & 0.341 & 0.175 & -0.166 & 80.0 & Entry/access \\
\midrule
\multicolumn{6}{l}{\textit{Diagnostics, not hard in the 14-moment rescore}}\\
Childless owner mean rooms & 6.224 & 6.0 & 6.053 & -0.171 & Implied by renter mean plus gap \\
Childless renter share rooms \(\ge 6\) & 0.138 & 25.0 & 0.100 & -0.038 & Renter large-unit tail \\
Old parent-childless ownership gap & 0.083 & 40.0 & 0.146 & 0.063 & Old family tenure diagnostic \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Current best fit.}
The best retained record under both the 17-moment loss and the 14-moment rescore
is DE wave 4, task 3, case 492. Its 14-moment loss is \(26.530\); its original
17-moment loss is \(26.903\). Dropping the three diagnostic moments lowers the
loss level slightly but does not change the selected basin.

\textbf{Identification map.}
Fertility level and childlessness discipline \(\kappa_n\),
\(\psi_{\mathrm{child}}\), and child costs. The two child-housing increments
discipline \(\bar h_{\mathrm{jump}}\) and \(\bar h_n\). Overall ownership,
young ownership, parent-child ownership gaps, and young liquid wealth discipline
entry/access and tenure parameters, especially \(b_0,\chi,\alpha,\beta\).
Renter rooms, the owner-renter room gap, and owner large-home use discipline the
space-demand and owner-ladder block. Old ownership and old wealth moments
discipline old retention, saving, and bequests, especially \(\beta,\theta_0\),
and \(\theta_n\). This is why removing a hard moment requires a replacement
moment for the same block.

\textbf{Main tension.}
The model does not need old households to exit ownership in an absolute sense.
The target is old ownership \(0.764\); the best model gives \(0.877\). At the
same time, young ownership should be \(0.341\), but the model gives \(0.175\),
and the owner-renter room gap should be \(2.419\), but the model gives \(1.493\).
The hard problem is joint: young entry is too low, old retention is too high,
and owner-renter space separation is too small.

The all-case frontier is sharper than the scalar loss. Among 263,282 retained
records, zero cases satisfy young ownership \(\ge 0.20\), old ownership
\(\le 0.90\), and room gap \(\ge 1.50\). Cases with old ownership and room gap
close to target usually have young ownership near zero. Cases with young
ownership close to target usually keep old ownership near one and compress the
room gap. This suggests the current model lacks a separating mechanism between
young entry, old retention, and large owner space.

\textbf{Next work.}
The 14-moment objective is cleaner and should be made explicit if it becomes
the diagnostic benchmark. But rescoring the existing search shows that this
alone does not fix the frontier. The next useful work is a focused verification
of the main frontier records, followed by a mechanism audit of tenure/access,
starter ownership versus large-owner housing, old downsizing/retention, and
old nonhousing wealth composition. Another blind global search is lower value
unless it tests a specific mechanism or an explicitly revised target system.

\end{document}
