% Review record for the quantification/calibration rewrite.
% This section was reviewed paragraph by paragraph and merged into the main paper
% after Tommaso's approval on 2026-07-17.
% Approval record as of 2026-07-17:
%   approved: calibration opening; Demographics and preferences;
%             Housing finance and transaction costs; Housing products and supply;
%             Entry wealth and bequests; Internally Estimated Parameters opening;
%             Saving; Housing demand, tenure, and supply;
%             Consumption needs and family size; Bequests.
%   provisionally approved, revisit: Income process.
%   superseded: the M4 footnote imposing kappa_T=0 (M5 estimates kappa_T).
%   approved and merged: Target Moments and Fit commentary.

\section{Calibration}
\label{sec:quant_calibration}

I organize the model parameters into two groups. The first consists of
parameters assigned from external evidence, measured directly in the data, or
fixed as maintained restrictions. The second consists of parameters estimated
internally by matching model moments to their empirical counterparts. I describe
the two groups in turn, explain which moments are particularly informative
about the internally estimated parameters, and then report the model's fit.

\subsection{Parameters Set Outside the Estimation}

I first describe the parameters assigned outside the estimation. Table~\ref{tab:quant_assigned}
summarizes their values and sources.

\paragraph{Demographics and preferences.}
A model period is four years. Households enter adult life at age 18, retire at
age 66, and live at most through age 85. Survival is certain before retirement
and thereafter follows four-year probabilities constructed from the combined
male and female survivor counts in the 2023 Social Security period life table.
Dependent children remain in the household for an expected 18 years, implying
a maturity probability of $2/9$ per period. I set risk aversion to $\sigma=2$.

\paragraph{Housing finance and transaction costs.}
Following \citet{greaneyDynamicUrbanEconomics2025}, I set the annual real
return on liquid wealth to $i_{\mathrm{ann}}=0.02$, annual housing depreciation
to $\delta_{\mathrm{ann}}=0.011$, the proportional selling cost to
$\psi^s=0.06$, and the labor-income tax rate to
$\tau_{\mathrm{pay}}=0.179$. The return is based on the average ten-year real
interest rate, depreciation combines BEA residential-structure depreciation
with an adjustment for the land share of house values, and the tax rate comes
from OECD data. I set the annual property-tax rate to
$\tau^p_{\mathrm{ann}}=0.01$. Mortgage debt cannot exceed a share $\phi=0.80$
of the value of the house, so a purchase requires at least 20 percent equity.
The benchmark imposes no separate payment-to-income limit.

\paragraph{Income process.}
I impose a deterministic lifecycle profile of household earnings, $e_a$,
normalized to average one over working ages. Within age, log income follows an
annual AR(1) process with persistence $\rho_z=0.960$, which I sample at the
model's four-year frequency and approximate with a five-state Rouwenhorst
chain. This persistence is near the upper end of the PSID-based estimates
summarized by \citet{sommerImplicationsUSTax2018}. I set the annual innovation
standard deviation to $\sigma_z=0.0645$, below estimates of gross income risk
in that literature. I treat this variance as provisional rather than as an
externally validated estimate of income risk.\footnote{With innovation risk in
the empirical range, households in the lowest persistent-income states cannot
always finance the model's minimum consumption and housing bundle. This
conflict may reflect missing insurance mechanisms---the current model has no
means-tested safety net, family transfers, or labor-supply margin---but it may
also indicate that the hard minimum requirements implied by the Stone--Geary
specification are too restrictive. Future revisions will therefore reconsider
both the missing insurance margins and the Stone--Geary preference structure.}
The quantitative results are conditional on this restriction.

\paragraph{Housing products and supply.}
Housing is indivisible and available in a limited assortment of types and
sizes \citep{davisHousingFinanceMacroeconomy2015}. I represent this lumpiness
with five owner-occupied house sizes,
$\{H_k\}_{k=1}^5=\{2,4,6,8,10\}$ rooms. Renters choose housing continuously up
to $\bar h^R=6$ rooms, the 90th percentile of the rental-room distribution in
the matched ACS sample. These menus summarize the observed tenure-specific
distribution of rooms. I set the
housing-supply elasticity to $\eta=1.75$, following the population-weighted
metropolitan estimate in \citet{saizGeographicDeterminantsHousing2010}.

\paragraph{Entry wealth and bequests.}
Households enter at age 18 with a draw from the PSID net-worth-to-income
distribution of childless renters aged 18--24. I impose $\theta_n=0$, so
bequest utility does not vary directly with the number of children. Among PSID
households aged 65--75, the median total-wealth-to-income ratio for households
with two or more children exceeds that for one-child households by $0.101$.
Children affect estate accumulation through their effects on housing needs,
tenure, and saving.

\begin{table}[H]
\centering
\caption{Parameters set outside the estimation}
\label{tab:quant_assigned}
\begingroup
\small
\renewcommand{\arraystretch}{1.05}
\begin{tabularx}{\textwidth}{@{}
>{\raggedright\arraybackslash}p{0.32\textwidth}
>{\raggedright\arraybackslash}p{0.22\textwidth}
>{\raggedright\arraybackslash}X@{}}
\toprule
Parameter & Value & Source or restriction \\
\midrule
Model period & 4 years & Model timing \\
Entry, retirement, and maximum ages & 18, 66, and 85 & Model timing \\
Post-retirement survival & Age-specific & 2023 Social Security period life table \\
Expected years with dependent children & 18 years & Child-maturity restriction \\
Risk aversion $\sigma$ & 2 & \\
Annual real return $i_{\mathrm{ann}}$ & 2.0\% &
Average 10-year real rate, 1962--2024; \citet{greaneyDynamicUrbanEconomics2025} \\
Annual housing depreciation $\delta_{\mathrm{ann}}$ & 1.1\% &
BEA structures depreciation, adjusted for the residential land share;
\citet{greaneyDynamicUrbanEconomics2025} \\
Annual property tax $\tau^p_{\mathrm{ann}}$ & 1.0\% & Maintained tax rate \\
Financed share $\phi$ & 0.80 &
20\% down-payment restriction; \citet{davisHousingFinanceMacroeconomy2015} \\
Sale cost $\psi^s$ & 6\% &
\citet{davisHousingFinanceMacroeconomy2015,greaneyDynamicUrbanEconomics2025} \\
Labor-income tax $\tau_{\mathrm{pay}}$ & 17.9\% &
OECD U.S. average; \citet{greaneyDynamicUrbanEconomics2025} \\
Income persistence $\rho_z$ & 0.960 annually & see text \\
Income innovation standard deviation $\sigma_z$ & 0.0645 annually & see text \\
Owner house sizes $\{H_k\}$ & $\{2,4,6,8,10\}$ rooms &
Occupied-room distribution in the ACS; see text \\
Largest rental unit $\bar h^R$ & 6 rooms & Matched ACS rental-size distribution \\
Housing-supply elasticity $\eta$ & 1.75 & \citet{saizGeographicDeterminantsHousing2010} \\
Entrant wealth distribution & Wealth-to-income distribution &
Childless renters aged 18--24 in the PSID \\
\bottomrule
\end{tabularx}
\endgroup
\end{table}

\subsection{Internally Estimated Parameters}

I estimate the remaining 14 parameters internally using 15 moments describing
completed fertility, childlessness, housing consumption, tenure, wealth
accumulation, and aggregate occupied housing. Table~\ref{tab:quant_parameters}
reports the estimates. Although the parameters are chosen jointly, I organize
the discussion into economic blocks and describe which moments are most
informative about each block.

\begin{table}[H]
\centering
\caption{Internally estimated parameters}
\label{tab:quant_parameters}
\small
\begin{tabularx}{0.84\textwidth}{@{}l X r@{}}
\toprule
Parameter & Interpretation & Estimate \\
\midrule
$\beta_{\mathrm{annual}}$ & Annual discount factor & 0.9912 \\
$\alpha_c$ & Consumption weight & 0.5912 \\
$\bar c_0$ & Baseline consumption requirement & 1.2596 \\
$\bar c_n$ & Consumption cost per dependent child & 0.3929 \\
$\bar h_0$ & Baseline housing requirement & 0.3915 \\
$\bar h_{\mathrm{jump}}$ & Housing-need increase with the first child & 1.6021 \\
$\bar h_n$ & Housing requirement per dependent child & 0.1644 \\
$\psi_{\mathrm{child}}$ & Utility from children & 0.1979 \\
$\kappa_F$ & Dispersion in family-size choices & 2.0434 \\
$\chi_O$ & Utility premium from owner-occupied housing & 1.1133 \\
$\bar H$ & Housing-supply scale & 8.6456 \\
$\theta_0$ & Strength of the bequest motive & 0.3118 \\
$\theta_1$ & Bequest curvature shift & 0.3973 \\
$\kappa_T$ & Dispersion in tenure choices & 0.0100 \\
\bottomrule
\end{tabularx}
\end{table}

\paragraph{Saving.}
The annual discount factor $\beta_{\mathrm{annual}}$ governs households'
willingness to carry resources across ages. Liquid wealth among young childless
renters, the older nonhousing-wealth share, and the late-life estate median are
particularly informative about it.

\paragraph{Housing demand, tenure, and supply.}
I use housing-size differences by family composition to discipline the housing
requirements. Mean rooms among childless renters is particularly informative
about $\bar h_0$, the increase associated with the first child disciplines
$\bar h_{\mathrm{jump}}$, and the difference between larger and smaller
families disciplines $\bar h_n$. The owner--renter rooms difference and the
share of childless owners in large homes help discipline $\alpha_c$ and
$\chi_O$. Ownership rates by age and family status are particularly informative
about $\chi_O$ and $\kappa_T$, while aggregate occupied rooms disciplines the
housing-supply scale $\bar H$.

\paragraph{Consumption needs and family size.}
The consumption requirement is $\bar c_0+\bar c_n n_s$. Outcomes for childless
households, including young liquid wealth, are particularly informative about
the baseline requirement $\bar c_0$, while fertility and differences in housing
and tenure by family composition help discipline the incremental cost
$\bar c_n$. Completed fertility and childlessness jointly discipline these
costs with the direct utility value of children $\psi_{\mathrm{child}}$ and the
dispersion of family-size choices $\kappa_F$. These parameters are estimated
jointly because the same family outcomes respond to both the cost and the
utility value of children.

\paragraph{Bequests.}
The bequest parameters $\theta_0$ and $\theta_1$ govern the strength and wealth
dependence of the bequest motive. The late-life median estate is particularly
informative about $\theta_0$, while the older nonhousing-wealth share, together
with old-age ownership, provides additional information about $\theta_1$.
These moments also respond to saving and tenure choices, so the two bequest
parameters are disciplined jointly with those blocks.

\subsection{Target Moments and Fit}

The target system combines completed fertility and childlessness from the CPS,
tenure and housing-size moments by age and family status from the ACS,
housing responses to changes in the number of children, and wealth moments
from the PSID. Table~\ref{tab:quant_fit} reports all 15 target moments and their
model counterparts.

\begin{table}[H]
\centering
\caption{Target moments and model fit}
\label{tab:quant_fit}
\small
\begin{tabularx}{0.94\textwidth}{@{}>{\raggedright\arraybackslash}X r r@{}}
\toprule
Moment & Data & Model \\
\midrule
\multicolumn{3}{@{}l}{\emph{Family size and family housing}} \\
Completed fertility & 1.918 & 2.034 \\
Childlessness rate & 0.188 & 0.189 \\
Family ownership gap & 0.168 & 0.219 \\
Housing increment with the first child, rooms & 0.664 & 0.681 \\
Rooms gap, 3$+$ versus 1--2 children, ages 30--55 & 0.368 & 0.607 \\
\addlinespace
\multicolumn{3}{@{}l}{\emph{Tenure and housing consumption}} \\
Ownership rate, ages 30--55 & 0.575 & 0.658 \\
Renter mean rooms, no children at home, ages 30--55 & 3.805 & 3.963 \\
Owner share with rooms $\ge6$, no children at home, ages 30--55 & 0.596 & 0.760 \\
Owner--renter mean rooms gap, no children at home, ages 30--55 & 2.419 & 2.323 \\
Ownership rate, ages 25--34 & 0.341 & 0.274 \\
Ownership rate, ages 65--75 & 0.764 & 0.954 \\
Mean occupied rooms, ages 18--85 & 5.780 & 5.673 \\
\addlinespace
\multicolumn{3}{@{}l}{\emph{Wealth and bequests}} \\
Young childless-renter liquid wealth / annual income, ages 25--35 & 0.179 & 0.328 \\
Median total estate wealth / annual income, ages 76--84 & 6.501 & 6.431 \\
Share with nonhousing wealth $\ge$ annual income, ages 65--75 & 0.608 & 0.610 \\
\bottomrule
\end{tabularx}
\end{table}

The model closely matches childlessness, the housing increment associated with
the first child, renter mean rooms, the owner--renter rooms difference,
aggregate occupied rooms, the late-life median estate, and the older
nonhousing-wealth share. It produces completed fertility of $2.034$, compared
with $1.918$ in the data. The model overstates the overall ownership rate, the
family ownership gap, the share of childless owners in large homes, and the
rooms difference
between larger and smaller families, while understating ownership at ages
25--34.

The wealth moments reveal a lifecycle tension. The two late-life wealth targets
are closely matched, but young childless renters accumulate liquid wealth equal
to $0.328$ of annual income, compared with $0.179$ in the PSID. The estimated
annual discount factor of $0.9912$ should therefore be read as a joint
calibration outcome rather than as a direct estimate of household patience.

The largest miss is old-age ownership: the model produces an ownership rate of
$0.954$ at ages 65--75, compared with $0.764$ in the ACS. The model has no
health or long-term-care expense risk and no age-dependent maintenance burden
that induces older owners to leave homeownership. Together with transaction
costs and a bequest motive that values the house at death, this makes owner
housing unusually persistent late in life. The untargeted estate distribution
shows a related limitation: the model produces an estate p90--p50 ratio of
$1.75$, compared with $3.45$ in the PSID, and understates nonhousing wealth in
the upper part of the late-life distribution. I therefore do not use the
current calibration to draw conclusions that depend on the late-life ownership
path or portfolio composition.

The calibration remains provisional because the income variance and the
late-life exit and portfolio margins require further work. The quantitative
exercises that follow focus on family housing needs, tenure, and the
down-payment constraint and should be read conditional on these limitations.
