\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,booktabs}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\setlength{\parskip}{3pt}
\title{\vspace{-1.4em}\textbf{Experiment E1: equivalence-scale preferences with sequential births}\\[2pt]
\large Design and calibration contract for the \texttt{intergen\_eqscale\_seq} fork}
\author{Fertility--housing project}
\date{July 18, 2026 --- an experiment; production and both prior forks untouched}
\begin{document}
\maketitle
\vspace{-1.4em}

\section*{1. What the experiment asks}

Two structural changes at once, each motivated by a measured failure of the
production model, recalibrated on the unchanged 15-moment target system so
the fit is comparable to M5. First, preferences lose their Stone--Geary
intercepts, so no income realization is ever infeasible and the income
process can carry literature-level risk ($\sigma_z=0.20$, set externally)
with no safety-net apparatus at all. Second, fertility becomes sequential:
households try for a first child, and one-child households may later try
for a second, each attempt succeeding with the age-declining fecundity
probability of the July-18 design note. The question ``take a look'' will
answer: can this architecture match the same data the production model
matches, and what do its timing margins and honest income risk do to the
objects the production model misses (the wealth tail, young liquid wealth,
the childlessness composition)?

\section*{2. Preferences: why not textbook Barten scales}

The obvious equivalence-scale implementation --- divide each good by a
child-dependent scale inside the existing composite --- is \emph{inert}
here: the period composite is Cobb--Douglas, so good-specific scales enter
log-additively and cancel from the allocation between consumption and
housing. Children would stop tilting housing demand entirely, removing the
paper's mechanism. The experiment therefore prices children through the two
devices that do work under Cobb--Douglas:
\[
u(c,h;n,s) \;=\; e(n,s)^{\,\sigma-1}\,
\frac{\bigl(c^{\alpha(n,s)}\,h^{1-\alpha(n,s)}\bigr)^{1-\sigma}}{1-\sigma}
\;+\;\psi\,n_k,
\]
with a \emph{share shifter}
$\alpha(n,s)=\operatorname{clip}\!\bigl(\alpha_0-(\delta_\alpha^{jump}+
\delta_\alpha n_k)\,\mathbf{1}\{\text{children present}\},\,0.05,\,0.95\bigr)$
--- children tilt expenditure toward housing, the demographic-taste-shifter
tradition of the consumption literature --- and a \emph{composite deflator}
$e(n,s)=1+\gamma_e n_k$, the utility cost of children (with $\sigma=2$ the
deflator multiplies the negative CRRA term, so larger families need more of
both goods to reach the same standard of living). There are no intercepts:
marginal utility diverges as consumption or housing services approach zero,
every strictly positive income draw admits a feasible choice, and the
empty-budget-set machinery (gates, censoring, floors) has nothing to catch.
The honest cost, stated once: the childless necessity structure is gone ---
poor childless households are no longer disciplined by a committed bundle
--- so the income and tenure gradients must be carried by the borrowing
constraint, prices, and the two new parameters. Whether they can be is part
of what the recalibration tests.

\section*{3. Fertility: the sequential margin}

As specified in the July-18 design note, with the v2 extension now active:
childless households choose \{wait, try for the first child\}; households
with one child and the window open choose \{stop, try for the second\};
attempts succeed with probability $\pi_j$ (external schedule, illustrative
until the demographic fit); a second birth lands in the existing
two-children raising state with the \emph{youngest-child clock restarted}.
The state grid does not grow: the collapse of per-child ages to one clock
--- the reason sequencing was originally dropped --- is retained, at the
approximation that the first child's maturation rides the youngest's
restarted clock (small against an 18-year expected duration). New measured
objects: first- and second-birth attempt and realized hazards by age, the
flow progression from one to two children, and the chosen-versus-clock
childlessness split. TFR and completed-fertility measurement are unchanged.

\section*{4. The E1 calibration contract}

Torch, 8 chains $\times$ 3:55 (the M5 template), smoke first. Targets: the
15 M5 moments, weights unchanged --- timing moments are reported
diagnostics, not targets, until their data counterparts are built.
Twelve free parameters, over-identified by construction:

\begin{center}\small
\begin{tabular}{lll}
\toprule
Parameter & Domain (transform) & Note \\
\midrule
$\beta$ (annual) & $[0.80, 0.9995]$ (discount) & external floor 0.94 still respected in reporting \\
$\alpha_0$ (\texttt{alpha\_cons}) & $[0.02, 0.98]$ (logit) & childless housing share \\
$\delta_\alpha^{jump}$ & $[0, 0.25]$ (softzero) & share tilt, any child \\
$\delta_\alpha$ & $[0, 0.25]$ (softzero) & share tilt per child \\
$\gamma_e$ & $[0, 3]$ (softzero) & composite deflator per child \\
$\psi$ (\texttt{psi\_child}) & $[-3, 3]$ (asinh) & child utility \\
$\kappa_f$ & $[0.02, 50]$ (log) & fertility logit scale \\
$\chi$ & $[0.10, 5]$ (log) & \\
$H_0$ & $[0.20, 80]$ (log) & \\
$\theta_0$ & $[0, 8]$ (softzero) & bequest \\
$\theta_1$ & $[0.02, 16]$ (log) & bequest luxury \\
$\kappa_T$ (\texttt{tenure\_choice\_kappa}) & $[0, 0.12]$ (softzero) & \\
\bottomrule
\end{tabular}
\end{center}

External restrictions (none free, none silent): $\theta_n=0$; $\sigma=2$;
$\phi=0.80$; income process $\rho_z=0.9602$, $\sigma_z=0.20$ (SS-range,
\emph{the experiment's point}); fecundity $(\omega_1,\omega_2,\bar a)=
(0.02, 0.134, 45)$ illustrative pending the demographic fit; no transfer
floor; the five Stone--Geary floor parameters do not exist in this
specification; entry wealth, survival, housing menus, taxes as in M5.
Seeds: M5 winner for shared coordinates; $(\delta_\alpha^{jump},
\delta_\alpha, \gamma_e) = (0.10, 0.05, 0.5)$.

Reading protocol for the result: report the full 15-row target-fit table
against M5's; the timing diagnostics; the estate p90/p50 and young liquid
wealth (the two objects honest risk moves in opposite directions in the
production model); and the childlessness composition. This is an
experiment: a failure to fit is itself the answer to a question worth
asking, and nothing in production moves either way.

\end{document}
